---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cmribio

<!-- badges: start -->
<!-- badges: end -->

This package contains functions and reporting templates used for standard analysis and reporting for bioinformatics and biostatistics projects. 

## Installation

You can install the development version of cmribio from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("jmvarberg/cmribio")
```

## Initialize Project

Each project should be contained in it's own project directory with specific subdirectories, and should contain a README file to capture any notes/information about the project to enable another member/user to continue working on the project if needed.

A helper function has been provided to create all subdirectories and a template README file.

```{r}
temp <- tempdir()
cmribio::init_proj(dir = temp, overwrite = TRUE)

```

We can see the subdirectories created:

```{r}
#| echo: false

fs::dir_tree(temp, recurse = TRUE)

```

A template README is also created, which you should open and modify as needed for your project.

```{r}
#| echo: false

txt_path <- file.path(temp, "README.txt")  # or some other file you've created
if (file.exists(txt_path)) {
  cat(readLines(txt_path), sep = "\n")
} else {
  message("File not found: ", txt_path)
}
```


## Pipeline Templates

Template files, functions and Quarto documents for automated reporting are provided for a few common workflows. These templates are created using the [`{targets}`](https://books.ropensci.org/targets/) package to provide scalable, efficient and automated pipelines and report generation. To see a list of the available templates, use the function 'cmribio::list_targets_templates()`:

```{r}
cmribio::list_targets_templates()
```

To use one of the templates, use the function `cmribio::use_target_template()` and provide one of the available values for 'Type' (returned by list_targets_templates()) to copy the pipeline and all associated files to the specified directory.

```{r}
temp <- tempdir()
cmribio::use_targets_template(dir = temp, type = "deseq2")
```


```{r}
#| echo: false

fs::dir_tree(temp, recurse = TRUE)

```
