---
title: "BCB Report"
subtitle: "CMRI Biostatistics and Computational Biology Core"
author: "Analyst"
date: today
format: 
  html:
    toc: true
    toc-depth: 5
    toc-location: left 
    toc-title: "Report Contents"
    embed-resources: false
    code-link: true
    code-fold: true
    code-tools:
        source: true
        toggle: true
    link-external-newwindow: true
    smooth-scroll: true
    theme: flatly
    
---

# Load Libraries

```{r}
library(tidyverse)
library(cowplot)
library(ggsci)
library(ggpubr)
library(rstatix)
library(data.table)
library(DT)
```


## Background and Objectives

Information on project background and analysis goals.


## Data Cleanup

Any processes for cleaning up/importing/formatting data.

## Data Analysis

### Section 1


### Section 2



### Section 3


## Results

### Embedded files

This can be used to directly embed results files, like CSVs or Excel files, into the document. It will then be self-contained, so anyone with the report can download the embedded file.

```{r, eval = F}
xfun::embed_file()
```



## Conclusions


## Session Info

```{r}
devtools::session_info()
```

