---
title: "Bulk RNAseq Analysis Report"
subtitle: "CMRI Biostatistics and Computational Biology Core"
author: "Last.First"
date: today
date-format: long
execute:
  echo: true
  eval: true
  warning: false
  error: false
  message: false
format:
  pdf:
    echo: false
    warning: false
    error: false
    message: false
  html:
    toc: true
    toc-depth: 5
    toc-location: left
    toc-title: "Report Contents"
    number-sections: false
    embed-resources: true
    code-link: true
    code-fold: true
    code-summary: "Show the code"
    code-tools:
        source: true
        toggle: true
    link-external-newwindow: true
    smooth-scroll: true
    theme: flatly
    
---

## Load Libraries

```{r}
#library(tidyverse)
#library(cowplot)
#library(ggsci)
#library(ggpubr)
#library(rstatix)
#library(data.table)
#library(DT)
library(DESeq2)
#library(plotly)
#library(Glimma2)
```


## Background and Objectives

Information on project background and analysis goals.

## Data Cleanup

Any processes for cleaning up/importing/formatting data.

## Data Import: nf-core DESeq object

For projects that used the nextflow nf-core/rnaseq analysis pipeline for upstream analysis, this option allows for import of the DESeq2 object that the pipeline generates as part of its QC. This object contains the following, based on alignment/quantification tools used:

* __STAR/RSEM:__ When STAR/RSEM is used for alignment and quantification, the DEseq2 object is built by default using the gene counts matrix generated by RSEM. This is located in the 'star_rsem/deseq2_qc' subdirectory of the nf-core/rnaseq output directory.

* __kallisto:__ 

In the chunk below, we show example code to read in the object, read in the samplesheet containing the sample metadata, modify if necessary, and then update the DESeq2 object. We also examine the DESeq2 object and make sure that sample names in the columns match samples in the metadata.

```{r deseq2 object import}
#| echo: true
#| eval: false #set this to true if you want to run using this approach

#Load the object created by nf-core/rnaseq
dds <- load("path/to/star_rsem/deseq2_qc/deseq2.dds.RData")

#read in the sample sheet used for the nf-core/rnaseq analysis
samplesheet <- data.table::fread("path/to/nfcore-rnaseq/samplesheet.csv")

#Modify as needed for your project here.
metadata_table <- samplesheet %>% 
  dplyr::select(sample, sample_title) %>% 
  dplyr::mutate(clean_names = janitor::make_clean_names(sample_title)) %>% 
  tidyr::separate_wider_delim(delim = "_", cols = clean_names, names = c("condition", "remove", "replicate")) %>% 
  dplyr::select(-remove) %>% 
  tibble::column_to_rownames(var = "sample")

#Set the colData for the DESeq object as the modified metadata table.
colData(dds) <- DataFrame(metadata_table)

all(colnames(counts(dds)) %in% rownames(colData(dds))) # Must be TRUE
 
all(colnames(counts(dds)) == rownames(colData(dds))) # Must be TRUE

dds <- DESeqDataSet(dds, design = ~ group)
 
# dds inspection
head(counts(dds)) # to check the raw counts
 
colData(dds) # to check the sample info
 
design(dds) # to check the design formula

```

## Data Import: Counts Matrix

An alternative approach is to read in the counts matrix and create a DESeq object. For example, running the nf-core/rnaseq pipeline with STAR/RSEM will generate mereged counts tables for both genes and counts. Here, we demonstrate how to read those in and build a DESeq object.

```{r build dds from counts}
gene_counts <- cmribio::converse_gene_counts
sample_sheet <- cmribio::converse_samplesheet

#fix the matrix to get gene IDs into the rownames
counts_matrix <- gene_counts %>% 
  dplyr::select(-`transcript_id(s)`) %>% 
  tibble::column_to_rownames(var = "gene_id") %>% 
  dplyr::mutate(across(where(is.numeric), as.integer))

#make the colData as needed for grouping and sampel information
metadata <- sample_sheet %>% 
  dplyr::select(sample, sample_description) %>% 
  tidyr::separate_wider_delim(sample_description, delim = " ", names = c("condition", "skip", "replicate")) %>%
  dplyr::mutate(condition = stringr::str_remove_all(condition, ",")) %>% 
  dplyr::select(-skip) %>% 
  DataFrame()

#Now, create DESeq object
dds <- DESeqDataSetFromMatrix(countData = counts_matrix, colData = metadata, design = ~ condition)

```

## QC Plots

First, let's make a boxplot of the number of reads in each sample.

### Reads per sample

```{r}

reads_plot <- function(dds, metadata, normalized = F) {

  if(!normalized) {
      #get the counts from the dds object
  counts <- counts(dds, normalized = F)
  title <- "Raw Counts"
  } else {
    counts <- counts(dds, normalized = T)
    title <- "Normalized Counts"
  }

  df <- data.frame(sample = colnames(counts),
                   total_counts = colSums(counts)) %>% 
    dplyr::left_join(as_tibble(metadata))
  # #pivot it long
  # long <- counts %>% 
  #   tibble::rownames_to_column(var = "geneid") %>% 
  #   tidyr::pivot_longer(cols = -geneid, names_to = "sample", values_to = "counts")
  
  #make plot
  ggplot(df, aes(x=sample, y = total_counts, fill = condition)) +
    geom_col() +
    ggsci::scale_fill_igv(name = "Condition") +
    xlab("Sample") +
    ylab("Total Reads") +
    coord_flip() +
    ggtitle(title) + 
    theme_bw()
  
}

reads_plot(dds, metadata)

```


## Data Analysis

### DESeq2 - Wald Test

Information here about what DESeq2 run with a Wald test specified does/is used for.

```{r}
dds <- DESeq2::DESeq(dds, test = "Wald")

```

### Normalization

Information about using rlog or vst for normalization and what those values should be used for.

```{r normalization}

rld <- rlog(dds, blind = TRUE)
vst <- vst(dds)
# 
# vst_counts <- assay(vst)
# rld_counts <- assay(rld)
# barplot(colSums(vst_counts))
# barplot(colSums(rld_counts))
```

### PCA Plot

Info about what a PCA plot is, how to intepret.

```{r pca plots}
pca_plot <- DESeq2::plotPCA(vst) + cowplot::theme_cowplot() + ggsci::scale_color_igv()
colData(rld)

pcaData <- plotPCA(vst, intgroup = "condition", returnData = TRUE)

ggplot(pcaData, aes(x= PC1, y = PC2, color = condition)) +
  geom_point() +
  theme_bw()
```


### Section 1


### Section 2



### Section 3


## Results

### Embedded files

This can be used to directly embed results files, like CSVs or Excel files, into the document. It will then be self-contained, so anyone with the report can download the embedded file.

```{r}
xfun::embed_file()
```


### Interactive Plots

For reports rendered to HTML, you may include interactive plots. Below, we show an example of how you can specify that the chunk with an interactive plot will only be rendered if the document is being knit to HTML - if it is knitting to a PDF or word document, then this section is skipped.

```{r .content-visible when-format="html"}




```



## Conclusions


## Session Info

```{r}
devtools::session_info()
```

