---
title: "DESeq2 Bulk RNAseq Analysis"
subtitle: "cmribio template report"
author: "<YOUR NAME>"
date: today
date-format: long
execute:
  echo: true
  eval: true
  warning: false
  error: false
  message: false
format: 
  html:
    toc: true
    toc-depth: 5
    toc-location: left 
    toc-title: "Report Contents"
    number-sections: true
    embed-resources: true
    code-link: true
    code-fold: true
    code-summary: "Show the code"
    code-tools:
        source: true
        toggle: true
    link-external-newwindow: true
    smooth-scroll: true
    theme: flatly
---

```{r}
library(targets)
library(tarchetypes)
library(cmribio)
library(tidyverse)
library(knitr)
library(cowplot)
library(PCAtools)
library(ggsci)
library(ggpubr)
library(DESeq2)
library(ComplexUpset)
library(UpSetR)
library(RColorBrewer)
library(pheatmap)
library(msigdbr)
library(limma)

```

# Overview and Summary

------------------------------------------------------------------------

List information here about the goals and objectives of the study, information on where data is located, how it has been pre-processed before analysis using the '_targets.R' bulk RNAseq DESeq2 analysis pipeline.

## Sample Sheet

Embed an interactive data table with all of the information that was included in the sample sheet.

```{r}
tar_read(sample_sheet_tracked) |> 
  cmribio::cmri_datatable()
```

## Column Data

Additional metadata was added to the samples for analysis and processing, which is captured in the column data here:



# Pipeline Status

```{r}
tar_visnetwork()
```

# PCA with PCAtools

Principle components analysis (PCA) 

```{r}
#| fig-width: 8
#| fig-height: 8

#PCAtools takes in the vst object from DESeq2. Load that from the targets pipeline.
targets::tar_load(vst_norm_object)

#Depending on how many design models you used, you'll need to access the correct one for plotting.
vst_obj <- assay(vst_norm_object[[1]])
vst_md <- colData(vst_norm_object[[1]])

#let's swap out with symbols
vst_obj <- cmribio::swap_ensmbl_for_symbols(vst_obj, species = "mouse")

p <- pca(vst_obj, metadata = vst_md, removeVar = 0.9) #use the top 10% most variable features by default.

screeplot(p, axisLabSize = 18, titleLabSize = 22)

biplot(p, colby = "group", colkey = c('Macrophage' = 'grey75', 'MNGC' = 'grey30', 'Stroma' = 'red'), labSize = 5, pointSize = 5)
cowplot::ggsave2("./results/plots/PCA_plot.pdf", height = 6, width = 6, units = "in")


#Make plot with top features in each PC loadings
biplot(p, showLoadings = TRUE,
    labSize = 5, pointSize = 5, sizeLoadingsNames = 5, ntopLoadings = 8)

```

# Sample-sample distance matrix with pheatmap

Following from the DESeq2 Vignette.

```{r}
#| fig-width: 14
#| fig-height: 12

#Get sample distances
sampleDists <- dist(t(vst_obj))
sampleDistMatrix <- as.matrix(sampleDists)

#Set color palette
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

#Set annotations
coldata <- as.data.frame(colData(vst_norm_object[[5]])) |> 
  dplyr::select(rpl, sex)

#set annotation colors for each column separately
ann_colors <- list(
  group = c(Macrophage = "navy", MNGC = "salmon", Stroma = "#E69F00")
)

pdf("./results/plots/Sample_sample_distance_matrix.pdf", height=6, width = 7)
pheatmap(sampleDistMatrix, 
         clustering_distance_rows = sampleDists, 
         clustering_distance_cols = sampleDists, 
         col = colors, 
         annotation_col = coldata,
         annotation_row = coldata,
         annotation_colors = ann_colors
)
dev.off()

```


# UpSet plot for UP/Down DEGs for each group 

```{r}
#| fig-width: 18
#| fig-height: 10

#read in the DE results object
tar_load(results_list)
de_results_list <- results_list[[1]]
names(de_results_list) <- c("RPL_vs_Control", "RPL003_vs_Control", "RPL003_vs_RPL", "R002_vs_Control", "R002_vs_RPL")

#get up significant hits
up_hits <- lapply(de_results_list, function(x) x |> dplyr::filter(padj <= 0.05, log2FoldChange > 0) |> tibble::rownames_to_column(var = "ensgene") |> dplyr::pull(ensgene))

down_hits <- lapply(de_results_list, function(x) x |> dplyr::filter(padj <= 0.05, log2FoldChange < 0) |> tibble::rownames_to_column(var = "ensgene") |> dplyr::pull(ensgene))

library(UpSetR)
library(ComplexUpset)

up_df <- UpSetR::fromList(up_hits) 
down_df <- UpSetR::fromList(down_hits)

up_plot <- ComplexUpset::upset(
  up_df,
  intersect = names(up_hits),    # <-- REQUIRED by ComplexUpset
  set_sizes = FALSE,  # nice default bars on the left
  n_intersections = 8, 
  base_annotations = list(
    'Intersection size' = intersection_size()
  )
) + 
  theme_cowplot() +
  ggtitle("Overlap of Up-regulated DEGs", subtitle = "Adj. p-val <= 0.05")

down_plot <- ComplexUpset::upset(
  down_df,
  intersect = names(down_hits),    # <-- REQUIRED by ComplexUpset
  set_sizes = FALSE,  # nice default bars on the left
  n_intersections = 8, 
  base_annotations = list(
    'Intersection size' = intersection_size()
  )
) + 
  theme_cowplot() +
  ggtitle("Overlap of Down-regulated DEGs", subtitle = "Adj. p-val <= 0.05")

cowplot::plot_grid(up_plot, down_plot)
ggsave2("./results/plots/UpSet_plots.pdf", height = 6, width = 12, units = "in")

```


# GSEA with limma::camera

Here, we run pathway enrichment using limma::cameraPR() for pre-ranked (log2FoldChange) values for each comparison.

```{r}
#| fig-width: 18
#| fig-height: 12

#get the quickomics DESeq2 test results
de_data <- data.table::fread("./results/Quickomics/group/quickomics_tests.csv") |> 
  dplyr::mutate(test = stringr::str_remove(test, "group_"),
                test = stringr::str_replace(test, "_", " vs "),
                test = as.factor(test))

#Get the gene sets of interest for pathway enrichment. Default to use Hallmark 50 only
genesets <- cmribio::get_genesets(species = "mouse", sets = "hallmark")

#run cameraPR using hte genesets chosen, on all comparisons in the DE data frame
camera_results <- cmribio::run_cameraPR(de_results = de_data, geneset_df = genesets, fc_col = "logFC", ens_col = "UniqueID", comp_col = "test")

#Combine into one output data frame
camera_res_comb <- camera_results |> 
  dplyr::bind_rows(.id = "Contrast") |> 
  dplyr::filter(FDR <= 0.01)

# write.csv(camera_res_hm_comb, "./results/Hallmark_camera_significant_FDR_0pt01.csv", row.names = F)

#Make a dot plot
camera_plots <- camera_dotplot(camera_res_comb, n_path = 10)

camera_plots$Up
camera_plots$Down

```

# Session Info

```{r}
sessionInfo()
```

