---
title: "DESeq2 Bulk RNAseq Analysis"
subtitle: "cmribio template report"
author: "<YOUR NAME>"
date: today
date-format: long
execute:
  echo: true
  eval: true
  warning: false
  error: false
  message: false
format: 
  html:
    toc: true
    toc-depth: 5
    toc-location: left 
    toc-title: "Report Contents"
    number-sections: true
    embed-resources: true
    code-link: true
    code-fold: true
    code-summary: "Show the code"
    code-tools:
        source: true
        toggle: true
    link-external-newwindow: true
    smooth-scroll: true
    theme: flatly
---

```{r}
library(targets)
library(tarchetypes)
library(cmribio)
library(tidyverse)
library(knitr)
library(cowplot)
library(PCAtools)
library(ggsci)
library(ggpubr)
library(DESeq2)
#library(ComplexUpset)
library(UpSetR)
library(RColorBrewer)
library(pheatmap)
library(msigdbr)
library(limma)
library(grid)

#Create output directories within results if they don't already exist
plots_out <- file.path(getwd(), "results/plots")
if(!dir.exists(plots_out)) {
  message("Creating plots output directory at /results/plots")
  dir.create(plots_out, recursive = TRUE)
}

```

# Overview and Summary

------------------------------------------------------------------------

List information here about the goals and objectives of the study, information on where data is located, how it has been pre-processed before analysis using the '_targets.R' bulk RNAseq DESeq2 analysis pipeline.

## Sample Sheet

Embed an interactive data table with all of the information that was included in the sample sheet.

```{r}
tar_read(sample_sheet_tracked) |> 
  cmribio::cmri_datatable()
```

## Column Data

Additional metadata was added to the samples for analysis and processing, which is captured in the column data here:



# Pipeline Status

```{r}
tar_visnetwork()
```

# PCA with PCAtools

Principle components analysis (PCA) 

```{r}
#| fig-width: 8
#| fig-height: 8

#PCAtools takes in the vst object from DESeq2. Load that from the targets pipeline.
targets::tar_load(vst_norm_object)

#Depending on how many design models you used, you'll need to access the correct one for plotting.
vst_obj <- assay(vst_norm_object[[1]])
vst_md <- colData(vst_norm_object[[1]])

#let's swap out with symbols
vst_obj <- cmribio::swap_ensmbl_for_symbols(vst_obj, species = "mouse")

p <- pca(vst_obj, metadata = vst_md, removeVar = 0.9) #use the top 10% most variable features by default.

screeplot(p, axisLabSize = 18, titleLabSize = 22)

biplot(p, colby = "group", colkey = c('Macrophage' = 'grey75', 'MNGC' = 'grey30', 'Stroma' = 'red'), labSize = 5, pointSize = 5, legendPosition = "bottom")
cowplot::ggsave2("./results/plots/PCA_plot.pdf", height = 6, width = 6, units = "in")


#Make plot with top features in each PC loadings
biplot(p, showLoadings = TRUE, colby = "group", colkey = c('Macrophage' = 'grey75', 'MNGC' = 'grey30', 'Stroma' = 'red'), labSize = 5, pointSize = 5, sizeLoadingsNames = 5, ntopLoadings = 8)

```

# Sample-sample distance matrix with pheatmap

Following from the DESeq2 Vignette.

```{r}
#| fig-width: 14
#| fig-height: 12

#Get sample distances
sampleDists <- dist(t(vst_obj))
sampleDistMatrix <- as.matrix(sampleDists)

#Set color palette
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

#Set annotations
coldata <- as.data.frame(colData(vst_norm_object[[1]])) |> 
  dplyr::select(sample, group)

#set annotation colors for each column separately
ann_colors <- list(
  group = c(Macrophage = "navy", MNGC = "salmon", Stroma = "#E69F00")
)

pdf("./results/plots/Sample_sample_distance_matrix.pdf", height=6, width = 7)
pheatmap(sampleDistMatrix, 
         clustering_distance_rows = sampleDists, 
         clustering_distance_cols = sampleDists, 
         col = colors, 
         annotation_col = coldata,
         annotation_row = coldata,
         annotation_colors = ann_colors
)
dev.off()

```

# Quickomics Files

Files were generated for uploading to [Quickomics](https://quickomics.bxgenomics.com/) for interactive data visualizations.

```{r}
tar_load(quickomics_files)

xfun::embed_dir("./results/Quickomics/")
```

# Metascape Files

We also generate files to upload to [Metascape](https://metascape.org/) for pathway and network analysis.

```{r}
tar_load(metascape_files)

xfun::embed_dir("./results/Metascape/")
```


# UpSet plot for UP/Down DEGs for each group 

```{r}
#| fig-width: 18
#| fig-height: 10

#read in the DE results object
tar_load(results_list)
de_results_list <- results_list[[1]]
comparisons <- unique(unlist(lapply(de_results_list, `[[`, "comparison")))
names(de_results_list) <- comparisons

#get up significant hits
up_hits <- lapply(de_results_list, function(x) x |> dplyr::filter(padj <= 0.05, log2FoldChange > 0) |> tibble::rownames_to_column(var = "ensgene") |> dplyr::pull(ensgene))

down_hits <- lapply(de_results_list, function(x) x |> dplyr::filter(padj <= 0.05, log2FoldChange < 0) |> tibble::rownames_to_column(var = "ensgene") |> dplyr::pull(ensgene))

up_df <- UpSetR::fromList(up_hits) 
down_df <- UpSetR::fromList(down_hits)

up_plot <- UpSetR::upset(up_df, order.by = "freq", nintersects = 10, group.by = "sets", text.scale = 1.8)
down_plot <- UpSetR::upset(down_df, order.by = "freq", nintersects = 10, group.by = "sets", text.scale = 1.8)
up_plot
grid.text("Up DEGs Overlap",x = 0.65, y=0.98, gp=gpar(fontsize=14))

down_plot
grid.text("Down DEGs Overlap",x = 0.65, y=0.98, gp=gpar(fontsize=14))


# up_plot <- ComplexUpset::upset(
#   up_df,
#   intersect = names(up_hits),    # <-- REQUIRED by ComplexUpset
#   set_sizes = FALSE,  # nice default bars on the left
#   n_intersections = 8, 
#   base_annotations = list(
#     'Intersection size' = intersection_size()
#   )
# ) + 
#   theme_cowplot() +
#   ggtitle("Overlap of Up-regulated DEGs", subtitle = "Adj. p-val <= 0.05")
# 
# down_plot <- ComplexUpset::upset(
#   down_df,
#   intersect = names(down_hits),    # <-- REQUIRED by ComplexUpset
#   set_sizes = FALSE,  # nice default bars on the left
#   n_intersections = 8, 
#   base_annotations = list(
#     'Intersection size' = intersection_size()
#   )
# ) + 
#   theme_cowplot() +
#   ggtitle("Overlap of Down-regulated DEGs", subtitle = "Adj. p-val <= 0.05")


#ggsave2("./results/plots/UpSet_plots.pdf", height = 6, width = 12, units = "in")

```


# GSEA with limma::camera

Here, we run pathway enrichment using limma::cameraPR() for pre-ranked (log2FoldChange) values for each comparison.

```{r}
#| fig-width: 18
#| fig-height: 12

#get the quickomics DESeq2 test results
de_data <- data.table::fread("./results/Quickomics/group/quickomics_comparison.csv") |> 
  dplyr::mutate(test = stringr::str_remove(test, "group_"),
                test = stringr::str_replace(test, "_", " vs "),
                test = as.factor(test))

#Get the gene sets of interest for pathway enrichment. Default to use Hallmark 50 only
genesets <- cmribio::get_genesets(species = "mouse", sets = "hallmark")

run_cameraPR_debug <- function(geneset_df, de_results,
                         fc_col = "log2FoldChange",
                         ens_col = "ensemble_gene_id",
                         comp_col = "test") {

    #split data frame of gene sets into a list
    gs_list <- split(x = geneset_df$ensembl_gene, f = geneset_df$gs_name)

    #convert dataframe with full DE results (not filtered for logFC or FDR) into list of named vectors with logFC values and ENSEMBL IDs as names
    #split by test
    de_list <- de_results |>
      dplyr::select(.data[[fc_col]], .data[[ens_col]], .data[[comp_col]]) |>
      jmvtools::named_group_split(.data[[comp_col]])

    #convert split into named vectors, values = logFC, names = ENSGENE
    de_list_vectors <- lapply(de_list, function(x) {
      x |> 
        dplyr::select(.data[[ens_col]], .data[[fc_col]]) |> 
        tibble::deframe()
    })

    #This function takes a list of gene sets, and a vector of logFC values with ENSGENE names. Creates and index and runs cameraPR on it.
    index_and_run <- function(gs_list, logFC_vector) {
      cam_index <- limma::ids2indices(gs_list, names(logFC_vector))
      camPR_res <- limma::cameraPR(statistic = logFC_vector, index = cam_index, use.ranks = TRUE)
    }

    #now, we run this on each logFC vector in our list of DE results - one for each comparison made
    camera_results <- lapply(de_list_vectors, index_and_run, gs_list = gs_list)

    return(camera_results)

}

camera_results <- run_cameraPR_debug(de_results = de_data, geneset_df = genesets, fc_col = "logFC", ens_col = "UniqueID", comp_col = "test")


# #run cameraPR using hte genesets chosen, on all comparisons in the DE data frame
# camera_results <- cmribio::run_cameraPR(de_results = de_data, geneset_df = genesets, fc_col = "logFC", ens_col = "UniqueID", comp_col = "test")

#Combine into one output data frame
camera_res_comb <- camera_results |> 
  dplyr::bind_rows(.id = "Contrast") |> 
  dplyr::filter(FDR <= 0.01) |> 
  tibble::rownames_to_column(var = "GeneSet")

# write.csv(camera_res_hm_comb, "./results/Hallmark_camera_significant_FDR_0pt01.csv", row.names = F)

#Make a dot plot
camera_plots <- camera_dotplot(camera_res_comb, n_path = 10)

camera_plots$Up
camera_plots$Down

```

# Session Info

```{r}
sessionInfo()
```

