---
title: "DESeq2 Bulk RNAseq Analysis"
subtitle: "cmribio template report"
author: "<YOUR NAME>"
date: today
date-format: long
execute:
  echo: true
  eval: true
  warning: false
  error: false
  message: false
format: 
  html:
    toc: true
    toc-depth: 5
    toc-location: left 
    toc-title: "Report Contents"
    number-sections: true
    embed-resources: true
    code-link: true
    code-fold: true
    code-summary: "Show the code"
    code-tools:
        source: true
        toggle: true
    link-external-newwindow: true
    smooth-scroll: true
    theme: flatly
---

```{r}
library(targets)
library(tarchetypes)
library(cmribio)
library(tidyverse)
library(knitr)
library(cowplot)
library(PCAtools)
library(ggsci)
library(ggpubr)
library(DESeq2)
library(ComplexUpset)
library(UpSetR)
library(RColorBrewer)
library(pheatmap)
library(msigdbr)
library(limma)

```

# Overview and Summary

------------------------------------------------------------------------

List information here about the goals and objectives of the study, information on where data is located, how it has been pre-processed before analysis using the '_targets.R' bulk RNAseq DESeq2 analysis pipeline.

## Sample Sheet

Embed an interactive data table with all of the information that was included in the sample sheet.

```{r}
tar_read(sample_sheet_tracked) |> 
  cmribio::cmri_datatable()
```

## Column Data

Additional metadata was added to the samples for analysis and processing, which is captured in the column data here:



# Pipeline Status

```{r}
tar_visnetwork()
```

# PCA with PCAtools

Principle components analysis (PCA) 

```{r}
#| fig-width: 8
#| fig-height: 8

#PCAtools takes in the vst object from DESeq2. Load that from the targets pipeline.
targets::tar_load(vst_norm_object)

#Depending on how many design models you used, you'll need to access the correct one for plotting.
vst_obj <- assay(vst_norm_object[[1]])
vst_md <- colData(vst_norm_object[[1]])

#let's swap out with symbols
vst_obj <- cmribio::swap_ensmbl_for_symbols(vst_obj, species = "mouse")

p <- pca(vst_obj, metadata = vst_md, removeVar = 0.9) #use the top 10% most variable features by default.

screeplot(p, axisLabSize = 18, titleLabSize = 22)

biplot(p, colby = "group", colkey = c('Macrophage' = 'grey75', 'MNGC' = 'grey30', 'Stroma' = 'red'), labSize = 5, pointSize = 5)
cowplot::ggsave2("./results/plots/PCA_plot.pdf", height = 6, width = 6, units = "in")


#Make plot with top features in each PC loadings
biplot(p, showLoadings = TRUE,
    labSize = 5, pointSize = 5, sizeLoadingsNames = 5, ntopLoadings = 8)

```

# Sample-sample distance matrix with pheatmap

Following from the DESeq2 Vignette.

```{r}
#| fig-width: 14
#| fig-height: 12

#Get sample distances
sampleDists <- dist(t(vst_obj))
sampleDistMatrix <- as.matrix(sampleDists)

#Set color palette
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

#Set annotations
coldata <- as.data.frame(colData(vst_norm_object[[5]])) |> 
  dplyr::select(rpl, sex)

#set annotation colors for each column separately
ann_colors <- list(
  group = c(Macrophage = "navy", MNGC = "salmon", Stroma = "#E69F00")
)

pdf("./results/plots/Sample_sample_distance_matrix.pdf", height=6, width = 7)
pheatmap(sampleDistMatrix, 
         clustering_distance_rows = sampleDists, 
         clustering_distance_cols = sampleDists, 
         col = colors, 
         annotation_col = coldata,
         annotation_row = coldata,
         annotation_colors = ann_colors
)
dev.off()

```


# UpSet plot for UP/Down DEGs for each group 

```{r}
#| fig-width: 18
#| fig-height: 10

#read in the DE results object
tar_load(results_list)
de_results_list <- results_list[[1]]
names(de_results_list) <- c("RPL_vs_Control", "RPL003_vs_Control", "RPL003_vs_RPL", "R002_vs_Control", "R002_vs_RPL")

#get up significant hits
up_hits <- lapply(de_results_list, function(x) x |> dplyr::filter(padj <= 0.05, log2FoldChange > 0) |> tibble::rownames_to_column(var = "ensgene") |> dplyr::pull(ensgene))

down_hits <- lapply(de_results_list, function(x) x |> dplyr::filter(padj <= 0.05, log2FoldChange < 0) |> tibble::rownames_to_column(var = "ensgene") |> dplyr::pull(ensgene))

library(UpSetR)
library(ComplexUpset)

up_df <- UpSetR::fromList(up_hits) 
down_df <- UpSetR::fromList(down_hits)

up_plot <- ComplexUpset::upset(
  up_df,
  intersect = names(up_hits),    # <-- REQUIRED by ComplexUpset
  set_sizes = FALSE,  # nice default bars on the left
  n_intersections = 8, 
  base_annotations = list(
    'Intersection size' = intersection_size()
  )
) + 
  theme_cowplot() +
  ggtitle("Overlap of Up-regulated DEGs", subtitle = "Adj. p-val <= 0.05")

down_plot <- ComplexUpset::upset(
  down_df,
  intersect = names(down_hits),    # <-- REQUIRED by ComplexUpset
  set_sizes = FALSE,  # nice default bars on the left
  n_intersections = 8, 
  base_annotations = list(
    'Intersection size' = intersection_size()
  )
) + 
  theme_cowplot() +
  ggtitle("Overlap of Down-regulated DEGs", subtitle = "Adj. p-val <= 0.05")

cowplot::plot_grid(up_plot, down_plot)
ggsave2("./results/plots/UpSet_plots.pdf", height = 6, width = 12, units = "in")

```


# GSEA with limma::camera

Here, we run pathway enrichment using limma::cameraPR() for pre-ranked (log2FoldChange) values for each comparison.

```{r}
#| fig-width: 18
#| fig-height: 12

#get the quickomics DESeq2 test results
de_data <- data.table::fread("./results/Quickomics/group/quickomics_tests.csv") |> 
  dplyr::mutate(test = stringr::str_remove(test, "group_"),
                test = stringr::str_replace(test, "_", " vs "),
                test = as.factor(test))

#split by test
de_list <- de_data |>
  dplyr::rename(ensembl_gene_id = UniqueID,
                log2FoldChange = logFC) |> 
  jmvtools::named_group_split(test)

#convert split into named vectors, values = logFC, names = ENSGENE
de_list_vectors <- lapply(de_list, tibble::deframe)

#get the geneset information from msigdbr
#Get all available gene set collections for humans
genesets_hs <- msigdbr_collections(db_species = "HS")
genesets_mm <- msigdbr_collections(db_species = "MM")
#pull the "C2" group which includes WikiPathways, KEGG, and others.
c2_genesets <- msigdbr(db_species = "HS", collection = "C2")

#Pull only the wikipathways subcollection
wp_genesets <- msigdbr(db_species = "HS", subcollection = "CP:WIKIPATHWAYS")

#subset the C2 gene sets to only keep gene sets that include senescence terms.
senescence_gs <- c2_genesets |> dplyr::filter(stringr::str_detect(gs_name, "SENESCENCE"))

#Pull the Hallmark 50 gene sets
hallmark_gs <- msigdbr(db_species = "HS", collection = "H")

#put the hallmark and senescence sets together
comb_gs <- dplyr::bind_rows(hallmark_gs, senescence_gs)

#put the hallmark plus full wikipathway gene sets together.
comb_hm_wp <- dplyr::bind_rows(hallmark_gs, wp_genesets)

#split into list by Gene set, with values of ENSGENES in each set

#HM + Senescence
gs_list <- split(x = comb_gs$ensembl_gene, f = comb_gs$gs_name)

#HM + Full WikiPathways
gs_list_full <- split(x = comb_hm_wp$ensembl_gene, f = comb_hm_wp$gs_name)

#Hallmark only
hm_list <- split(x = hallmark_gs$ensembl_gene, f = hallmark_gs$gs_name)

#now, need to create camera index
run_camera <- function(gs_list, stat_vector) {
  
  #make index
  cam_index <- limma::ids2indices(gs_list, names(stat_vector))
  
  #run cameraPR
  cam_res <- limma::cameraPR(statistic = stat_vector, index = cam_index, use.ranks = TRUE)
}

#run the cameraPR on all de_list_vectors

#Run camera on Hallmark only
camera_res_full <- lapply(de_list_vectors, run_camera, gs_list = hm_list)
camera_res_full <- lapply(camera_res_full, function(x) x |> tibble::rownames_to_column(var = "GeneSet"))

#Combine into one output data frame
camera_res_hm_comb <- camera_res_full |> 
  dplyr::bind_rows(.id = "Contrast") |> 
  dplyr::filter(FDR <= 0.01)

write.csv(camera_res_hm_comb, "./results/GSEA_Camera/Combined_Hallmark_GSEA_with_camera_significant_FDR_0pt01_sex_karyotype_rpl.csv", row.names = F)

#Run camera on Hallmark + Full WikiPathways
camera_hm_wp <- lapply(de_list_vectors, run_camera, gs_list = gs_list_full)
camera_hm_wp <- lapply(camera_hm_wp, function(x) x |> tibble::rownames_to_column(var = "GeneSet"))

#Combine into one output data frame
camera_res_hm_wp_comb <- camera_hm_wp |> 
  dplyr::bind_rows(.id = "Contrast") |> 
  dplyr::filter(FDR <= 0.01)

write.csv(camera_res_hm_wp_comb, "./results/GSEA_Camera/Combined_Hallmark_plus_Full_WikiPathways_GSEA_with_camera_significant_FDR_0pt01_sex_karyotype_rpl.csv", row.names = F)

#Run camera on Hallmark + 23 Senescence Pathways
camera_res_full_sen <- lapply(de_list_vectors, run_camera, gs_list = gs_list)
camera_res_full_sen <- lapply(camera_res_full_sen, function(x) x |> tibble::rownames_to_column(var = "GeneSet"))

#Combine into one output data frame
camera_res_hm_comb_sen <- camera_res_full_sen |> 
  dplyr::bind_rows(.id = "Contrast") |> 
  dplyr::filter(FDR <= 0.01)

write.csv(camera_res_hm_comb_sen, "./results/GSEA_Camera/Combined_Hallmark_plus_senescence_GSEA_with_camera_significant_FDR_0pt01_sex_karyotype_rpl.csv", row.names = F)

#Make some plots
camera_plot <- function(x, n_path = 10) {
  
  #Check that Contrast is a factor
  if(is.factor(x$Contrast) == FALSE) {
    print("Contrast column in input data frame not passed as a factor. Coercing to factor with default level names. Change levels in input and re-run if different levels are desired.")
    x$Contrast <- factor(x$Contrast, levels = unique(x$Contrast))
  }

  #Specifying the color mapping to levels in the contrast column (which is a factor)
  contrast_lvls <- levels(x$Contrast)
  n_colors <- length(contrast_lvls)
  
  if(n_colors <= 10) {
    #npg palette handles up to 10 colors
    contrast_cols <- ggsci::pal_npg("nrc")(n_colors)
  } else {
    #igv palette handles up to 51 colors
    contrast_cols <- ggsci::pal_igv("default")(n_colors)
  }
  
  names(contrast_cols) <- contrast_lvls

  #get the range of values for NGenes for Up and Down
  ext_fun    <- scales::breaks_pretty(n = 4)
  size_breaks <- ext_fun(c(min(x$NGenes), max(x$NGenes)))

  #modify the pathway labels
  x <- x |> 
    dplyr::mutate(GeneSet = stringr::str_replace_all(GeneSet, "_", " ")) |> 
                  #GeneSet = stringr::str_to_sentence(GeneSet))
    dplyr::group_by(Direction, GeneSet) |> 
    dplyr::mutate(MaxSig = max(-log10(FDR)))
  
  #Get top 10 pathways Up and Down
  top_gs <- x |> dplyr::ungroup() |> 
    dplyr::group_by(Direction, GeneSet) |> 
    dplyr::summarise(MaxSigFilt = max(MaxSig)) |> 
    dplyr::ungroup() |> 
    dplyr::group_by(Direction) |> 
    dplyr::arrange(desc(MaxSigFilt)) |> 
    dplyr::slice_head(n=n_path) |> 
    jmvtools::named_group_split(Direction)
  #want dot plot with x = FDR, size = NGenes
  up <- x |> 
    dplyr::filter(Direction == "Up", GeneSet %in% top_gs$Up$GeneSet) |> 
    ggplot(aes(x=-log10(FDR), y = reorder(GeneSet, MaxSig), size = NGenes, fill = Contrast)) +
    geom_point(alpha = 0.7, color = "black", pch = 21) +
    scale_size_area(
      name   = "# Genes in Set",
        breaks = size_breaks,
        limits = c(min(size_breaks), max(size_breaks)),
        max_size = 10,
        oob = scales::oob_squish
    ) +
    facet_wrap(~Direction) +
    ylab("") +
    scale_fill_manual(values = contrast_cols) +
    theme_bw() +
    theme(
      strip.background = element_blank(),
      strip.text = element_text(face = "bold", color = "black", size = 12)
  ) +
    guides(fill = guide_legend(override.aes = list(size = 6), order = 1))
  
  down <- x |> 
    dplyr::filter(Direction == "Down", GeneSet %in% top_gs$Down$GeneSet) |> 
    ggplot(aes(x=-log10(FDR), y = reorder(GeneSet, MaxSig), size = NGenes, fill = Contrast)) +
    geom_point(alpha = 0.7, color = "black", pch = 21) +
    scale_size_area(
      name   = "# Genes in Set",
        breaks = size_breaks,
        limits = c(min(size_breaks), max(size_breaks)),
        max_size = 10,
        oob = scales::oob_squish
    ) +
    facet_wrap(~Direction) +
    ylab("") +
    scale_fill_manual(values = contrast_cols) +
    theme_bw() +
    theme(
      strip.background = element_blank(),
      strip.text = element_text(face = "bold", color = "black", size = 12)
  ) +
    guides(fill = guide_legend(override.aes = list(size = 6), order = 1))
  
  #combine plots into one output plot
  out <- cowplot::plot_grid(up, down, ncol = 1)

  return(out)

}

camera_res_hm_comb <- camera_res_hm_comb |> dplyr::mutate(Contrast = factor(Contrast, levels = c("R002 vs Control", "R002 vs RPL", "R003 vs Control", "R003 vs RPL", "RPL vs Control")))
camera_plots <- camera_plot(camera_res_hm_comb)
camera_plots
cowplot::ggsave2("./results/plots/Camera_pathway_dotplots_karyotype_rpl.pdf", height = 8, width = 10, units = "in")

camera_plots_sen <- camera_plot(camera_res_hm_comb_sen, n_path = 15)
camera_plots_sen
cowplot::ggsave2("./results/plots/Camera_pathway_dotplots_karyotype_rpl_with_senescence.pdf", height = 10, width = 12, units = "in")


camera_plots_hm_wp <- camera_plot(camera_res_hm_wp_comb, n_path = 15)
camera_plots_hm_wp
cowplot::ggsave2("./results/plots/Camera_pathway_dotplots_karyotype_rpl_hallmark_wikipathways.pdf", height = 10, width = 12, units = "in")
```

# Heat Maps For Pathways of Interest
```{r}
#| fig-width: 10
#| fig-height: 16
library(DESeq2)
targets::tar_load(vst_norm_object)
#we want the 5th object in the list, which is ~ sex + group, with RPL003, RPL and Control as groups
vst_obj <- assay(vst_norm_object[[6]])

#expectation: expression matrix with ENSG row IDs (human), input pathway name(s) from msigdbr, and list from msigdbr named by pathway with values of ENS IDs
pathway_heatmaps <- function(expr_mat, gs_list, pathway_name, show_rowNames = FALSE) {

  # expr_mat <- vst_obj
  # gs_list = hm_list
  # pathway_name = "HALLMARK_MYC_TARGETS_V2"
  # 
  #get genes to plot
  goi <- unique(gs_list[[pathway_name]])
  
  expr <- expr_mat[rownames(expr_mat) %in% goi, ]
  
  #now, swap out ENSGENE IDs with Symbols
  ids <- rownames(expr)

  grch38 <- annotables::grch38 |> 
    dplyr::mutate(symbol = dplyr::if_else(symbol == "", ensgene, symbol)) |> 
    dplyr::distinct(ensgene, .keep_all = TRUE)

  swap <-  data.frame(ensgene = ids) |> 
    dplyr::left_join(grch38) |> 
    dplyr::mutate(symbol = make.unique(symbol))
  
  #swap out the row names for symbols
  rownames(expr) <- swap$symbol
  
  #make heatmap
  if(show_rowNames) {
    pheatmap::pheatmap(expr, 
                     cluster_rows = TRUE, 
                     cluster_cols = TRUE, 
                     scale = "row", 
                     color = colorRampPalette(c("darkblue", "white", "firebrick"))(100), 
                     show_rownames = TRUE)
  } else {
    pheatmap::pheatmap(expr, 
                     cluster_rows = TRUE, 
                     cluster_cols = TRUE, 
                     scale = "row", 
                     color = colorRampPalette(c("darkblue", "white", "firebrick"))(100), 
                     show_rownames = FALSE)
    
    
  }
  
  
}

myc_v1_hm <- pathway_heatmaps(expr_mat = vst_obj, gs_list = hm_list, pathway_name = "HALLMARK_MYC_TARGETS_V1")

myc_v2_hm <- pathway_heatmaps(expr_mat = vst_obj, gs_list = hm_list, pathway_name = "HALLMARK_MYC_TARGETS_V2", show_rowNames=TRUE)

pdf("./results/plots/HALLMARK_MYC_TARGETS_V2_heatmap.pdf", height = 10, width = 8)
myc_v2_hm
dev.off()

tgfb_hm <- pathway_heatmaps(expr_mat = vst_obj, gs_list = hm_list, pathway_name = "HALLMARK_TGF_BETA_SIGNALING", show_rowNames=TRUE)

pdf("./results/plots/HALLMARK_TGF_BETA_SIGNALING_heatmap.pdf", height = 10, width = 8)
tgfb_hm
dev.off()
```



# Session Info

```{r}
sessionInfo()
```

